---
title: "Explore One Variable"
author: "Giorgos Traskas"
date: "2017-06-02"
output: html_document
---

## Reading in Data
```{r}
# getwd()
list.files()
pf <- read.csv("pseudo_facebook.tsv", sep = "\t")
names(pf)
```

## Load ggplot and ggthemes
```{r}
library(ggplot2)
library(ggthemes)
```

## Histogram of Users' Birthdays
```{r}
theme_set(theme_minimal(10))
ggplot(aes(x = dob_day), data = pf) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)
```

## Friend Count
```{r}
ggplot(data = pf, aes(x = friend_count)) +
  geom_histogram()
```

## Limiting the Axes
```{r}
ggplot(data = pf, aes(x = friend_count)) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1000))
```

## Adjusting Binwidth and Friend Count by Gender
```{r}
ggplot(data = pf, aes(x = friend_count)) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

## Omitting NA Gender Observations
```{r}
ggplot(data = subset(pf, !is.na(gender)), aes(x = friend_count)) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

## Omitting Any NA Observations
```{r}
ggplot(data = na.omit(pf), aes(x = friend_count)) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

## Statistics by Gender
```{r}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

## Tenure in Days
```{r}
ggplot(aes(x = tenure), data = pf) + 
   geom_histogram(binwidth = 30, color = 'black', fill = '#099DD9')
```

## Tenure in Years and Labeling Plots
```{r}
ggplot(aes(x = tenure/365), data = pf) + 
   geom_histogram(binwidth = .25, color = 'black', fill = '#F79420') +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7)) + 
  xlab('Number of years using Facebook') + 
  ylab('Number of users in sample')
```

## User Ages
```{r}
ggplot(data = pf, aes(x = age)) +
  geom_histogram(binwidth = 1, color = 'black', fill = '#F79420') +
  scale_x_continuous(breaks = seq(0, 113, 5))
```

## Load gridExtra
```{r}
library(gridExtra)
```

## Transforming Data
```{r}
summary(pf$friend_count)
summary(log10(pf$friend_count + 1))
summary(sqrt(pf$friend_count))
```

## Define Individual Plots and Arrange in Grid
```{r}
p1 <- ggplot(data = pf, aes(x = friend_count)) + geom_histogram()
p2 <- ggplot(data = pf, aes(x = sqrt(friend_count))) + geom_histogram()
p3 <- ggplot(data = pf, aes(x = log10(friend_count + 1))) + geom_histogram()
p4 <- p1 + scale_x_sqrt()
p5 <- p1 + scale_x_log10()

grid.arrange(p1, p2, p3, p4, p5, ncol = 1)
```

## Frequency Polygons
```{r}
p6 <- ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') +
  ylab('Percentage of users with that friend count')

p7 <- ggplot(aes(x = friend_count, y = ..density..), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') +
  ylab('Percentage of users within each group')

grid.arrange(p6, p7, ncol = 2)
```

## Likes on the Web
```{r}
ggplot(aes(x = www_likes), data = subset(pf, !is.na(gender))) + 
  geom_freqpoly(aes(color = gender)) + 
  scale_x_log10()
```

## How many www_likes there are in the entire data set for males?
```{r}
by(pf$www_likes, pf$gender, sum)
```

## Box Plots
```{r}
ggplot(aes(x = gender, y = friend_count), data = subset(pf, !is.na(gender))) + 
  geom_boxplot()
```

## Limiting Box Plots
```{r}
p8 <- ggplot(aes(x = gender, y = friend_count), data = subset(pf, !is.na(gender))) + 
  geom_boxplot()

p9 <- ggplot(aes(x = gender, y = friend_count), data = subset(pf, !is.na(gender))) + 
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 1000))

p10 <- ggplot(aes(x = gender, y = friend_count), data = subset(pf, !is.na(gender))) + 
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 1000))

grid.arrange(p8, p9, p10, ncol = 1)
```

## Friend Count by Gender
```{r}
by(pf$friend_count, pf$gender, summary)
```

## Who made more friend requests on average - men or women?
```{r}
by(pf$friendships_initiated, pf$gender, summary)
```

## Getting Logical
```{r}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0)

pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)
```

## What percent of check in using mobile?
```{r}
sum(pf$mobile_check_in == 1)/length(pf$mobile_check_in)
```

## Load Diamonds Dataset
```{r}
data("diamonds")
summary(diamonds)
str(diamonds)
nrow(diamonds)
names(diamonds)
dim(diamonds)
sapply(diamonds, class)
?diamonds
```

## Histogram of Price
```{r}
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram()
```

## Price Statistics
```{r}
summary(diamonds$price)
```

## Diamond Cost Counts
```{r}
subset(diamonds, price < 500)
subset(diamonds, price < 250)
subset(diamonds, price >= 15000)
length(which(diamonds$price < 500))
length(which(diamonds$price < 250))
length(which(diamonds$price >= 15000))
library(plyr)
count(diamonds, 'cut')
```

## Cheaper Diamonds
```{r}
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 1, color = "orange") +
  scale_x_continuous(breaks = seq(0, 2000, 100), limits = c(0, 2000))
```

## Price by Cut Histograms
```{r}
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 1, color = "brown") +
  facet_wrap(~cut)
```

## Price by Cut
```{r}
by(diamonds$price, diamonds$cut, max)
by(diamonds$price, diamonds$cut, min)
by(diamonds$price, diamonds$cut, median)
ggplot(aes(x = cut, y = price), data = diamonds) + 
  geom_boxplot(color = "brown")
```

## Scales and Multiple Histograms
```{r}
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 1, color = "brown") +
  facet_wrap(~cut, scales="free")
```

## Price per Carat (i.e. price/carat) by Cut
```{r}
ggplot(data = diamonds, aes(x = price/carat)) +
  geom_histogram(binwidth = 0.1, color = 'white', fill = "orange") +
  facet_wrap(~cut, scales="free_y") +
  scale_x_log10()
```

## Price per Carat Box Plots
```{r}
p11 = ggplot(data = diamonds, aes(x = color, y = price/carat, fill = color)) +
  geom_boxplot() +
  stat_summary(fun.y=mean, geom="point", shape=1, size=3) +
  scale_fill_brewer(palette="Spectral")
p12 = ggplot(data = diamonds, aes(x = cut, y = price/carat, fill = cut)) +
  geom_boxplot() +
  stat_summary(fun.y=mean, geom="point", shape=1, size=3) +
  scale_fill_brewer(palette="Set2")
p13 = ggplot(data = diamonds, aes(x = clarity, y = price/carat, fill = clarity)) +
  geom_boxplot() +
  stat_summary(fun.y=mean, geom="point", shape=1, size=3) +
  scale_fill_brewer(palette="Set3")
grid.arrange(p11, p12, p13, ncol = 1)
```

## Interquartile Range - IQR
```{r}
by(diamonds$price, diamonds$color, summary)
IQR(subset(diamonds, color == "D")$price)
IQR(subset(diamonds, color == "J")$price)
```

## Price per Carat Box Plots by Color and for Ideal and IF Clarity
```{r}
ideal_IF = subset(diamonds, cut == "Ideal" & clarity == "IF")
ggplot(data = ideal_IF, aes(x = color, y = price/carat, fill = color)) +
  geom_boxplot() +
  stat_summary(fun.y=mean, geom="point", shape=1, size=3) +
  scale_fill_brewer(palette="Spectral")
```

## Carat Frequency Polygon
```{r}
ggplot(aes(x = carat), data = diamonds) +
  geom_freqpoly(binwidth=0.01) +
  scale_x_continuous(limits = c(0.2, 1.2), breaks=seq(0,6,0.1))
```
