---
title: "White Wines Exploration"
author: "George Traskas"
date: "2017-06-26"
output: 
  html_document:
    code_folding: hide
    toc: true
    theme: spacelab
---
========================================================

# Description of Dataset and Preparation

The dataset which is related to white variants of the Portuguese "Vinho Verde"
wine, consists of several physicochemical sample test values and an output
sensory variable. The output value is the median of at least 3 evaluations made
by wine experts. Each expert graded the wine quality between 0 (very bad)
and 10 (very excellent).

Variables info:

1. Fixed acidity: most acids involved with wine or fixed or nonvolatile (do not
evaporate readily).

2. Volatile acidity: the amount of acetic acid in wine, which at too high of
levels can lead to an unpleasant, vinegar taste.

3. Citric acid: found in small quantities, citric acid can add 'freshness' and
flavour to wines.

4. Residual sugar: the amount of sugar remaining after fermentation stops,
it's rare to find wines with less than 1 g/L and wines with greater than 45 g/L
are considered sweet.

5. Chlorides: the amount of salt in the wine.

6. Free sulphur dioxide: the free form of SO2 exists in equilibrium between
molecular SO2 (as a dissolved gas) and bisulphite ion; it prevents microbial
growth and the oxidation of wine.

7. Total sulphur dioxide: amount of free and bound forms of S02; in low
concentrations, SO2 is mostly undetectable in wine, but at free SO2
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine.

8. Density: the density of water is close to that of water depending on the
percent alcohol and sugar content.

9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic)
to 14 (very basic); most wines are between 3-4 on the pH scale.

10. Sulphates: a wine additive which can contribute to sulphur dioxide gas
(S02) levels, which acts as an antimicrobial and antioxidant.

11. Alcohol: the percent alcohol content of the wine.

12. Quality: output variable based on sensory data with values between 0 and 10.

### Load Libraries and Data

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# # Install packages if necessary
# install.packages("pander")
# install.packages("ggplot2")
# install.packages("gridExtra")
# install.packages("plotly")
# install.packages("scales")
# install.packages("memisc")
# install.packages("tidyr")
# install.packages("formattable")
# install.packages("highcharter")
# install.packages("DT")
# install.packages("corrplot")
# install.packages("devtools")
# devtools::install_github("hrbrmstr/ggalt")
# install.packages("rpart")
# install.packages("rpart.plot")
# install.packages("DBI")

# Load packages
library(pander)
library(ggplot2)
library(gridExtra)
library(plotly)
library(shiny)
library(scales)
library(memisc)
library(tidyr)
library(formattable)
library(corrplot)
library(ggalt)
library(rpart)
library(rpart.plot)
```

```{r Load_View_Data}
# Load Data
wines <- read.csv("wineQualityWhites.csv")

# View data
formattable(wines, list(area(col = c(quality)) 
                        ~ normalize_bar("yellow", 0.2))) %>% as.datatable()
```

# Univariate Plots Section

Take a look of the data structure and get a statistical summary.

```{r results='asis', summary}
# Create a new bound form of SO2 variable.
wines$bound.sulfur.dioxide <- wines$total.sulfur.dioxide -
  wines$free.sulfur.dioxide

# Create total acidity.
wines$total.acidity <- wines$fixed.acidity +
  wines$volatile.acidity

# Create a rating variable from quality.
wines$rating <- NA
wines$rating <- ifelse(wines$quality < 5, "Undrinkable",
                       ifelse(wines$quality < 6, "Drinkable",
                              ifelse(wines$quality < 7, "Average",
                                     ifelse(wines$quality < 8, "Good", "Great"))))
wines$rating <- factor(wines$rating)
wines$rating <- ordered(wines$rating, levels = c("Undrinkable", "Drinkable",
                                                 "Average", "Good", "Great"))

# str(wines) # uncomment to see the structure of wines dataset.
pandoc.table(summary(wines))
```

The white wines dataset includes 12 variables with almost 5000 observations.
Residual sugar concentration is too low for most of the wines with a maximum
value at about 66 g/L and 3rd quantile at 9.9. Therefore, there are only few
sweet white wines in the list, since sweetness considered for values greater
than 45 g/L.

For most wines free sulphur dioxide is below 50 ppm. Values of pH ranges
from 2.7 to 4 and alcohol content ranges from 8 to 14. Quality ranges only
from 3 to 9 (no 0, 1, 2 and 10 values) with a median at 6.

The following histograms will show the distribution of the data.
Boxplots will help to depict the outlies, although in the following analysis
all of the data were used and no outliers were removed. Since the data were
derived from chemical tests, it can be safely considered that for each
observation at least 2 measurements were conducted in order to handle
experimental errors. As it is known, an outlier may be due to variability in
the measurement or it may indicate experimental error; the latter only is
sometimes excluded from the dataset.

```{r Univariate_Plots_1}
# Create a function for plotting.
f_plot1 <- function(x, bins) {
  grid.arrange(ggplot(data = wines, aes_string(x = 1, y = x)) +
                 geom_jitter(alpha = 0.05) +
                 geom_boxplot(alpha = 0.2, color = "blue"),
               ggplot(data = wines, aes_string( x = x)) +
                 geom_histogram(bins = bins), ncol = 2)
}

f_plot1("total.acidity", 40)
```

Distribution of total acidity is normal with few values less than 4 and a long
tail after 9. Most values are observed at 7-8.

```{r Univariate_Plots_2}
f_plot1("citric.acid", 35)
```

It seems that citric acid has some 0 values and few outliers after 0.6. The 
interquartile range (IQR) is from 0.27 to 0.39.

```{r Univariate_Plots_3}
f_plot1("residual.sugar", 50)
```

There is a positive skewness for residual sugar. The transformation of x axis
to the log10 scale, as well as the tuning of the bins may reveal something.

```{r Univariate_Plots_3_1}
ggplot(data = wines, aes( x = residual.sugar)) +
  geom_histogram(bins = 40) +
  scale_x_log10()
```

Indeed, residual sugar distribution is bimodal with peaks at about 2.5 and 
around 10.

```{r Univariate_Plots_4}
f_plot1("chlorides", 35)
```

Some tweaking is necessary to help visualise the histogram of chlorides.

```{r Univariate_Plots_4_1}
ggplot(data = wines, aes( x = chlorides)) +
  geom_histogram(bins = 35) +
  scale_x_continuous(limit = c(0, 0.1), breaks = seq(0, 0.1, 0.01))
```

Chlorides follow a normal distribution with 50% of values from 0.036 to 0.05.

```{r Univariate_Plots_5}
f_plot1("total.sulfur.dioxide", 40)
```

Total sulphur dioxide depicts a wide range from 9 to 440. An x axis
transformation to the log10 scale may show more information.

```{r Univariate_Plots_5_1}
ggplot(data = wines, aes( x = total.sulfur.dioxide)) +
  geom_histogram(bins = 60) +
  scale_x_log10()
```

It looks like a negative skewed distribution, after the log10 scale conversion.

```{r Univariate_Plots_6}
f_plot1("density", 30)
```

Density has a small range with an IQR from 0.9917 to 0.9961.
Only very few wines have a density higher than 1 with a maximum at 1.0390.

```{r Univariate_Plots_7}
f_plot1("pH", 55)
```

The distribution of pH is normal with values from 2.720 to 3.820 and a median
at 3.18, which shows that wines are acidic.

```{r Univariate_Plots_8}
f_plot1("sulphates", 22)
```

Sulphates' distribution is positively skewed with a median at 0.47.

```{r Univariate_Plots_9}
f_plot1("alcohol", 21)
```

Alcohol's histogram seems to be positively skewed with values from 8 to 14.20.

```{r quality}

p1 <- ggplot(data = wines, aes(x = rating)) + 
  geom_bar()

p2 <- ggplot(data = wines, aes(x = quality)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  scale_x_continuous(breaks = seq(3, 9, 1))

grid.arrange(p1, p2, ncol = 1)
```

Wines with high residual sugar concentrations probably have the higher density,
as sugars contribute to density. Checking sweetness,

```{r message=FALSE, warning=FALSE, sweet_wines}
by(wines$density, wines$residual.sugar > 45, summary)
table(wines$residual.sugar > 45)
```

it is found that there is only one sweet wine, which also has the higher
observed density in the list.

# Univariate Analysis

### What is the structure of your dataset?

* The white wines dataset consists of 4898 observations with 11 numerical
variables and one variable, quality, which can be considered as categorical.
* Zero numerical values were observed only for the citric acid variable.
* Residual sugar is the only variable which follows a bimodal distribution.
This clearly shows that there are two groups of wines:

    * Dry wines with low sugars below 5 g/L
    * Semi or Off-Dry wines with sugars higher than 5 g/L (probably late
    harvest wines)

### What is/are the main feature(s) of interest in your dataset?

Although quality is considered the most important feature, the relationships
between all of the variables will be further studied.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

It is known that pH, acidity, sugar content, SO2 concentrations and salinity
(chlorides) play an important role in the taste, flavours, aromas, structure,
colour and ageability of wines.

### Did you create any new variables from existing variables in the dataset?

Three new variables were created:

* Bound sulphur dioxide which is the difference between total and free SO2.
* Total acidity variable from the addition of fixed and volatile acidity,
since its overall value determines how a wine will taste, feel in mouth and how
well it will age.
* A new rating categorical variable based on the existing quality grade.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

* Residual sugar concentration was the only bimodal distribution which clearly
shows two types of white wines in the list.
* A new rating quality parameter was created as an orderd factor variable to
facilitate and improve the exploratory analysis.
* Histograms' bin size was tuned and in some cases the x scale was transformed
to log10 in order to reveal new patterns in the data.

# Bivariate Plots Section

Explore the correlations between the variables.

```{r fig.width=10, fig.height=10, correlations}
wines_corr_r <- round(cor(wines[c(2:13)]), 1)
par(xpd=TRUE)
corrplot(wines_corr_r, method = "number", mar = c(2, 0, 1, 0))
```

Several interesting correlations were found:

* Strong correlations
    * Positive
        * sugars - density, as it was expected
    * Negative
        * density - alcohol, since alcohol in water tends to decrease density

* Weaker correlations
    * Positive
        * citric acid - fixed acidity
        * density - fixed acidity
        * sugars - free SO2
        * sugars - total SO2
        * chlorides - density
        * free SO2 - density
        * total SO2 - density
        * quality - alcohol
    * Negative
        * sugars - alcohol, since during fermentation sugars are converted
        into alcohol
        * pH - fixed acidity
        * chlorides - alcohol
        * total SO2 - alcohol

Examine further the above relationships visualizing them with scatterplots and
encircling wines with quality over 9.

```{r fig.width=12, fig.height=10, Bivariate_Plots_Strong}

p1 <- ggplot(data = wines, aes(x = alcohol, y = rating)) +
  geom_jitter(alpha = 0.1)

p2 <- ggplot(data = wines, aes(x = round(alcohol / 0.5) * 0.5, y = quality)) +
  geom_line(stat = "summary", fun.y = mean)

p3 <- ggplot(data = wines, aes(x = residual.sugar, y = density)) +
  geom_jitter(alpha = 0.05) +
  xlim(0, 20) +
  ylim(0.98, 1.01) +
  geom_smooth(method = "lm", color = "blue") +
  geom_encircle(aes(x = residual.sugar, y = density),
                data = wines[wines$quality >= 9, ],
                color = "orangered")

p4 <- ggplot(data = wines, aes(x = alcohol, y = density)) +
  geom_jitter(alpha = 0.05) +
  ylim(0.98, 1.01) +
  geom_smooth(method = "lm", color = "blue") +
  geom_encircle(aes(x = alcohol, y = density),
                data = wines[wines$quality >= 9, ],
                color = "orangered")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

At first glance, no visualising trend was revealed for quality-alcohol, even
after applying jitter and changing transparency to prevent overplotting.
However, the plot of alcohol vs. mean quality shows an increase in quality with
alcohol number. Density-sugars pair exhibits a strong positive (+0.8) linear
relationship and density-alcohol presents a negative one (-0.8).

```{r Bivariate_Plots_Weak_1}
p1 <- ggplot(data = wines, aes(x = chlorides, y = density)) +
  geom_jitter(alpha = 0.05) +
  xlim(0.01, 0.07) +
  ylim(0.985, 1.005) +
  geom_smooth(method = "lm", color = "blue")

p2 <- ggplot(data = wines, aes(x = total.sulfur.dioxide, y = density)) +
  geom_jitter(alpha = 0.05) +
  xlim(0, 300) +
  ylim(0.985, 1.005) +
  geom_smooth(method = "lm", color = "blue")

grid.arrange(p1, p2)
```

Salts content, i.e. chlorides and sulphates increase the density of the wine as 
sugars do too, but this time the correlation is weaker (+0.3).

```{r Bivariate_Plots_Weak_2}
ggplot(data = wines, aes(x = residual.sugar, y = alcohol)) +
  geom_jitter(alpha = 0.1) +
  xlim(0, 20) +
  geom_smooth(method = "loess", color = "blue")
```

Residual sugar is negatively connected (-0.5) to alcohol, since it is the source
of ethanol during winemaking.

```{r Bivariate_Plots_Weak_3}
ggplot(data = wines, aes(x = pH, y = fixed.acidity)) +
  geom_jitter(alpha = 0.1) +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "blue")
```

Generally, the lower the pH, the higher the acidity in the wine, which was also
confirmed by a moderate negative correlation (-0.4).

```{r Bivariate_Plots_Weak_4}
ggplot(data = wines, aes(x = alcohol, y = total.sulfur.dioxide)) +
  geom_jitter(alpha = 0.1) +
  ylim(0, 300) +
  geom_smooth(method = "lm", color = "blue")
```

Sulphur dioxide has some degree of inhibitory affect on yeast during
fermentation (convertion of sugars to alcohol) and this may be a good reason of
alcohol reduction with the increase of SO2.

```{r Bivariate_Plots_Weak_5}
ggplot(data = wines, aes(x = alcohol, y = chlorides)) +
  geom_jitter(alpha = 0.1) +
  ylim(0, 0.1) +
  geom_smooth(method = "lm", color = "blue")
```

Chloride salts like SO2, present similar retarding effects on the activity of
the yeast. That is the reason for the negative correlation -0.4.

```{r Bivariate_Boxplots_1}

# Create a function for plotting.
f_plot2 <- function(x, y) {
  ggplot(wines, aes_string(x = "rating", y = y)) +
  geom_boxplot(alpha = 0.5,  color = "blue") +
  geom_jitter(alpha = 0.02) +
  stat_summary(fun.y = mean,
               geom = "point",
               shape = 8,
               size = 3,
               color = "blue") +
  theme(legend.position = "none") +
  theme(axis.title.x = element_blank())
}

f_plot2(y = "total.acidity") + ylim(5, 10)
```

The higher total acidity is observed for undrinkable wines. Let's see if there
is any trend in the scatter plot.

```{r Scatterplot_acid_quality}
ggplot(data = wines, aes(x = total.acidity, y = quality)) +
  geom_jitter(alpha = 0.1) +
  xlim(5, 10) +
  geom_smooth(method = "lm")
```

A very small decrease of quality with an increase in acidity is revealed.

```{r Bivariate_Boxplots_2}
f_plot2(y = "chlorides") + ylim(0, 0.1)
```

Undrinkable and drinkable wines have the higher chlorides. It is known that
increased chloride concentrations give an undesirable soapy taste in wine.

```{r Scatterplot_chlorides_quality}
ggplot(data = wines, aes(x = chlorides, y = quality)) +
  geom_jitter(alpha = 0.1) +
  xlim(0, 0.1) +
  geom_smooth(method = "lm")
```

The more the chlorides the lower the quality.

```{r Bivariate_Boxplots_3}
p1 <- f_plot2(y = "free.sulfur.dioxide") + ylim(0, 100)

p2 <- f_plot2(y = "total.sulfur.dioxide") + ylim(0, 300)

grid.arrange(p1, p2, ncol = 2)
```

Undrinkable wines distinguish with the lowest free and total SO2. Probably these
wines start to deteriorate, since the lack of SO2 favours unwanted microbial
growth and oxidations.

```{r Scatterplot_SO2_quality}
ggplot(data = wines, aes(x = free.sulfur.dioxide, y = quality)) +
  geom_jitter(alpha = 0.1) +
  xlim(0, 100) +
  geom_smooth(method = "lm")
```

Unfortunately, the study of the scatterplots of the above free SO2-quality pair
reveals no visual trends.

```{r fig.width=12, fig.height=14, Bivariate_Boxplots_4}
p1 <- f_plot2(y = "citric.acid") + ylim(0, 0.75)

p2 <- f_plot2(y = "residual.sugar") + ylim(0, 20)

p3 <- f_plot2(y = "density") + ylim(0.99, 1)

p4 <- f_plot2(y = "pH")

p5 <- f_plot2(y = "sulphates") + ylim(0.2, 1)

p6 <- f_plot2(y = "alcohol")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)
```

It can be clearly seen the unusual trend of alcohol-quality pair
(positive correlation +0.4) and therefore, the negative density-quality
relation (-0.3), since alcohol and density are competitive factors (-0.8).

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The main feature of interest, quality, seems to vary with alcohol, density, SO2,
sugars and chlorides. Generally, a "bad" wine in this list has a higher acidity,
chlorides and density, while its SO2 levels are lower.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

It is worth noting the weak negative correlation of SO2-alcohol. A better 
understanding of the chemistry of SO2 can explain a competitive relation between
SO2 and alcohol.

Alcohol is produced by the Saccharomyces yeast strains during alcoholic
fermentation. Sulphur dioxide (SO2) is a natural by-product of winemaking as
a small quantity is produced during the alcoholic fermentation by yeasts.

In practice however, SO2 is added by the winemakers either as a preservative or
prior alcoholic fermentation to control the growth of microorganisms, which 
stops the production of alcohol. This benefits the flavours of white wines,
since the enzyme polyphenol oxidase is inhibited and less oxidative browning of
the juice occurs. This helps to preserve the fruity and floral aromas found in
the juice.

### What was the strongest relationship you found?

The strongest relationships was between density and sugars or alcohol:

* sugars in wine tend to increase density, and
* alcohol decrease density.

# Multivariate Plots Section

Classify wines to dry and off-dry type based on the residual sugar content.
Create a new data frame, that contains information on quality and type of wine.

```{r fig.width=5, fig.height=4, new_dataframe_lineplot}
# Create a type variable from residual.sugar.
wines$type <- NA
wines$type <- ifelse(wines$residual.sugar < 5,
                     "dry", "off_dry")

wines.alcohol_by_type <- wines %>% 
  group_by(quality, type) %>% 
  summarise(mean_alcohol = mean(alcohol),
            median_alcohol = median(alcohol),
            n = n()) %>% 
  arrange(quality)

ggplot(data = wines.alcohol_by_type, 
       aes(x = quality, y = mean_alcohol)) +
  geom_line(aes(color = type)) +
  scale_x_continuous(breaks = seq(3, 9, 1))
```

There is an increasing trend in quality with alcohol which is more important for
the dry wines but obviously more data are needed to make robust conclusions.

```{r fig.width=12, fig.height=14, Multivariate_Plots}

f_plot3 <- function(x, y) {
  ggplot(data = wines, 
  aes_string(x = x, y = y, color = "rating")) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type='seq', guide = guide_legend(title = "Rating"))
}

p1 <- f_plot3(x = "residual.sugar", y = "density") +
  xlim(0, 25) + ylim(0.985, 1.005)

p2 <- f_plot3(x = "alcohol", y = "density") +
  ylim(0.985, 1.005)

p3 <- f_plot3(x = "chlorides", y = "density") +
  ylim(0.985, 1.005) + xlim(0, 0.2)

p4 <- f_plot3(x = "total.sulfur.dioxide", y = "density") +
  ylim(0.985, 1.005) + xlim(0, 300)

p5 <- f_plot3(x = "residual.sugar", y = "alcohol") +
  xlim(0, 20)

p6 <- f_plot3(x = "pH", y = "fixed.acidity") +
  ylim(4, 10)

p7 <- f_plot3(x = "alcohol", y = "chlorides") +
  ylim(0, 0.15)

p8 <- f_plot3(x = "alcohol", y = "total.sulfur.dioxide") +
  ylim(0, 300)

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol = 2)
```

The above plots depict a different visual behaviour ("the separation") between
lower and higher quality wines. For example, "good" and "great" wines have 
a higher increase of density than "average" wines due to chlorides content.

A first attempt to build a linear model and use the variables in the linear
model to predict the quality of a wine was not very successful, since the
biggest R-squared value is only 0.282.

```{r lm}
m1 <- lm(quality ~ alcohol, data = wines)
m2 <- update(m1, ~ . + density)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + chlorides)
m5 <- update(m4, ~ . + total.sulfur.dioxide)
m6 <- update(m5, ~ . + fixed.acidity)
m7 <- update(m6, ~ . + residual.sugar)
m8 <- update(m7, ~ . + pH)
m9 <- update(m8, ~ . + sulphates)
m10 <- update(m9, ~ . + free.sulfur.dioxide)
m11 <- update(m10, ~ . + citric.acid)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, sdigits = 3)
```

The second attempt was to build a decision tree model, since it is more robust,
handles non linearity and performs well with both numerical and categorical
data. Two types of the wine dataset were examined:

* The dataset as is:

```{r decision_tree_model1}
fit <- rpart(rating ~ alcohol + density + volatile.acidity + chlorides +
               total.sulfur.dioxide + fixed.acidity + residual.sugar +
               pH + sulphates + free.sulfur.dioxide + citric.acid,
               data = wines,
               method="class")

rpart.plot(fit)
```

* The dataset with some variables transformed to categorical:

```{r warning=FALSE, message=FALSE, decision_tree_model2}
# Create new categorical variables with high-low levels.
wines$free.sulfur.dioxide_hilo <- NA
wines$free.sulfur.dioxide_hilo <- ifelse(wines$free.sulfur.dioxide < 50, 0, 1)
wines$free.sulfur.dioxide_hilo <- factor(wines$free.sulfur.dioxide_hilo)

wines$chlorides_hilo <- NA
wines$chlorides_hilo <- ifelse(wines$chlorides < 0.06, 0, 1)
wines$chlorides_hilo <- factor(wines$chlorides_hilo)

wines$volatile.acidity_hilo <- NA
wines$volatile.acidity_hilo <- ifelse(wines$volatile.acidity < 0.26, 0, 1)
wines$volatile.acidity_hilo <- factor(wines$volatile.acidity_hilo)

wines$citric.acid_hilo <- NA
wines$citric.acid_hilo <- ifelse(wines$citric.acid > 0, 1, 0)
wines$citric.acid_hilo <- factor(wines$citric.acid_hilo)

fit <- rpart(rating ~ alcohol +
               density +
               volatile.acidity_hilo +
               chlorides_hilo +
               free.sulfur.dioxide_hilo +
               fixed.acidity +
               residual.sugar +
               pH +
               sulphates +
               total.sulfur.dioxide +
               citric.acid_hilo,
               data = wines,
               method="class")

rpart.plot(fit)
```

Few of the variables seems to play a significant role in quality, such as
alcohol, volatile acidity, free SO2 and sugars.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Several interesting relations were found and most of them were explained by
literature reviewing. Some of the most important correlation pairs are:

* Density vs sugars, alcohol and chlorides
* Quality vs alcohol
* Alcohol vs sugars

### Were there any interesting or surprising interactions between features?

It was really surprising the quite strong positive relationship of quality and
alcohol, since ethanol is not considered tasty.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Two models, a linear and a decision tree model were created to predict the
quality of a wine, but the results were unsatisfactory. Obviously, additional
observations or variables are needed to build a good prediction model.

------

# Final Plots and Summary

### Plot One
```{r fig.width=10, fig.height=8, Plot_One}
p1 <- ggplot(data = wines, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 0.06, colour = "white",
                 alpha = 0.8, aes(y = ..density.., fill = ..count..)) +
  scale_fill_gradient("Count", low = "white", high = "lightblue") +
  scale_x_log10() +
  geom_density(colour = "lightblue") +
  theme_classic() +
  labs(title = "Histogram and Density Plot of Residual Sugar in White Wines",
       x = "Residual Sugar in g/L",
       y = "Density")

ggplotly(p1)
```

### Description One

The distribution of the residual sugar appears to be bimodal on log scale,
because there are two types of wine, the dry and the off-dry ones.

### Plot Two
```{r fig.width=10, fig.height=8, Plot_Two}
p2 <- plot_ly(data = wines[wines$density < 1.002, ], 
              x = ~density) %>%
  add_markers(y = ~alcohol,
              name = 'Alcohol',
              marker = list(color = 'rgba(88, 116, 152, 0.9)'),
              hoverinfo = "text",
              text = ~paste(alcohol)) %>%
  add_lines(y = ~fitted(lm(alcohol ~ density)),
            line = list(color = '#E86850'),
            name = "Linear smoother", 
            showlegend = FALSE) %>%
  add_markers(y = ~residual.sugar, 
              name = 'Residual Sugar',
              yaxis = 'y2', 
              hoverinfo = "text", 
              alpha = 0.5, 
              text = ~paste(residual.sugar, 'g/L')) %>%
  add_lines(y = ~fitted(loess(residual.sugar ~ density)),
            line = list(color = '#FFD800'),
            name = "Loess Smoother", 
            showlegend = TRUE) %>%
  layout(title = "How Alcohol and Residual Sugar Relate with Density",
         xaxis = list(title = "Density"),
         yaxis = list(side = 'left', 
                      title = "Percentage of Alcohol Content",
                      showgrid = FALSE, 
                      zeroline = FALSE),
         yaxis2 = list(side = 'right', 
                       overlaying = "y", 
                       title = "Residual Sugar in g/L", 
                       showgrid = FALSE, 
                       zeroline = FALSE))

ggplotly(p2)
```

### Description Two

Alcohol in wines tend to decrease their density in contrast to sugars which
increases the density. Actually, it seems to be a linear correlation between
alcohol and density with R-squared at 61%.

```{r}
summary(lm(I(density) ~ I(alcohol), data = wines))
```

### Plot Three
```{r fig.width=14, fig.height=8, Plot_Three}
wines_good <- subset(wines, wines$free.sulfur.dioxide < 50 & 
                          wines$volatile.acidity < 0.26 & 
                          wines$citric.acid > 0 &
                          wines$chlorides < 0.06 &
                          wines$sulphates < 0.47)

p3 <- ggplot(data = wines_good, 
  aes(x = alcohol, y = density, color = rating)) +
  geom_point(size = 2, alpha = 0.9) +
  geom_smooth(method = "lm", se = FALSE, size = 0.8) +
  scale_color_brewer(type='seq', guide = guide_legend(title = "Rating")) +
  scale_x_continuous(limits = c(8.5, 14),
                     breaks = seq(8.5, 14, 0.5)) +
  scale_y_continuous(limits = c(0.985, 1.001),
                     breaks = seq(0.985, 1.001, 0.002)) +
  ggtitle("Alcohol vs. Density
          with low SO2, Chlorides, Volatile Acidity
          and high Citric Acid Aromas") +
  labs(x = "Percentage of Alcohol Content",
       y = "Density (g/mL)") +
  theme_classic()

ggplotly(p3)
```

### Description Three

This plot shows again how density varies with alcohol number but this time only
for selected wines with the following criteria:

* free sulphur dioxide < 50 ppm (over this threshold SO2 becomes evident in the
nose and taste of wine),
* sulphates < 0.47 (contribute to sulphur dioxide gas levels),
* chlorides < 0.06 g/L (increased levels of sodium chloride appears to convey
undesirable soapy notes),
* volatile acidity < 0.26 (high levels can lead to an unpleasant, vinegar
taste),
* citric acid > 0 (add 'freshness' and flavour to wines).

This was an attempt to examine wines with desirable physicochemical properties,
but unfortunately this bucket still contains all qualities. Nevertheless,
density decreases with alcohol.

------

# Reflection

The white wines data set contains information on almost 5000 wines across 12
variables. Several interesting trends and relations were observed during the
data exploration. There was a clear bimodal distribution of residual sugar which
indicates two types of wines, dry and off-dry. The quality ranged from 5 to 7
for most of wines. There were strong correlations of density-sugars and
density-alcohol, as it was expected, and an unusual positive relation of
quality and alcohol. Several other relations were also investigated.

Eventually, a linear and a decision tree method were applied in order to build a
prediction model of the quality using all the variables of the data set. It was
found once again that alcohol presents the higher influence in wine quality, but
the significance of the rest factors was not the desirable. Obviously, this
data set is too small or there is a need for other more crucial variables to be
investigated.

It is really difficult to tightly define with only 12 physicochemical properties
the complexity of a wine, which can come from different things such as:

* the vineyard and the notion of terroir (the set of all environmental factors
that affect a crop's phenotype, unique environment contexts and farming
practices),
* winemaking techniques (the more artisan and non-industrial the approach the
more complex the resulting wine), and
* pre and post bottling maturation.

------

# References

* http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html
* https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
* https://en.wikipedia.org/wiki/Sugars_in_wine
* http://www.winespectator.com/display/show/id/scoring-scale
* https://www.thespruce.com/which-wine-has-the-least-sugar-3511283
* http://winemakersacademy.com/understanding-wine-acidity/
* http://winemakersacademy.com/?s=sulfur
* http://thewinehub.com/home/2013/01/09/the-use-or-not-of-sulfur-dioxide-in-winemaking-trick-or-treat/
* https://www.practicalwinery.com/janfeb09/page4.htm
* http://www.winespectator.com/drvinny/show/id/51772
* http://winemakersacademy.com/?s=ph
* http://winemakersacademy.com/?s=sulfites
* http://www.decanter.com/features/alcohol-levels-the-balancing-act-246426/
* https://www.scientificamerican.com/article/salty-science-floating-eggs-in-water/
* http://waterhouse.ucdavis.edu/whats-in-wine/fixed-acidity
* https://www.practicalwinery.com/janfeb09/page3.htm
* http://www.reluctantgourmet.com/salt-does-kill-yeast/
* http://novyden.blogspot.gr/2013/07/ggplot-inside-function-needs-some-care.html#
* https://www.analyticsvidhya.com/blog/2016/04/complete-tutorial-tree-based-modeling-scratch-in-python/
* http://trevorstephens.com/kaggle-titanic-tutorial/r-part-3-decision-trees/
* https://rstudio-pubs-static.s3.amazonaws.com/123908_3df6d25c8ad04030a2a429661125a59b.html#/3
* http://wineoscope.com/2015/10/02/when-a-wine-is-salty-and-why-it-shouldnt-be/
* http://www.decanter.com/learn/wine-terminology/sulfites-in-wine-friend-or-foe-295931/
* https://vinepair.com/wine-blog/why-low-sulfite-wines-taste-better/
* http://www.thekitchn.com/wine-words-complexity-169017
* https://en.wikipedia.org/wiki/Terroir
