---
title: "Compute Statistics from Card Draws"
author: "George Traskas"
date: "2017-07-04"
output: 
  html_document:
    code_folding: hide
    toc: true
    theme: spacelab
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

# Source
# https://jefflirion.github.io/udacity/Intro_to_Descriptive_Statistics/Project.html
```

## Distribution of Card Values

```{r histogram}
# Load libraries.
library(ggplot2)


# Create the deck.
faces <- c("Ace", as.character(seq(2, 10)), "Jack", "Queen", "King")

deck <- data.frame(cards = c(paste0(faces, " of spades"), 
                             paste0(faces, " of hearts"), 
                             paste0(faces, " of diamonds"), 
                             paste0(faces, " of clubs")), 
                   values = rep(c(seq(1, 9), rep(10, times = 4)), times = 4), 
                   stringsAsFactors = FALSE)

# Create the histogram.
ggplot(data = deck, aes(x = values)) + 
  geom_histogram(binwidth = 1, alpha = 0.5, fill = "cyan", colour = "white") + 
  labs(x = "Value", y = "Count", title = "Card Value Histogram") +
  scale_x_continuous(breaks = seq(1, 10), limits=c(0.5, 10.5)) +
  scale_y_continuous(breaks = seq(0, 16, 4))
```

* What is the mean of the card values for a single draw?
* What is the median of the card values for a single draw?
* What is the standard deviation of the card values for a single draw?
(population sd)

```{r statistics}
summary(deck)

# Find the population sd.
sqrt(mean((deck$values-mean(deck$values))^2))
```

## Get Samples from the Deck of Cards

To obtain a single sample, shuffle your deck of cards and draw three cards from 
it. (You will be sampling from the deck without replacement.) Record the cards 
that you have drawn and the sum of the three cards??? values. Repeat this sampling
procedure a total of at least thirty times.

```{r get_samples}
# The number of cards to pick
m <- 3

# The number of times to sample the deck
n <- 10000

# Initialize the data frame and name the columns.
samples <- data.frame(matrix(nrow = n, ncol = m + 1))
colnames(samples) <- c(paste0("card", seq(1:m)), "sum")

# Sample the deck n times, drawing m cards.
set.seed(1)
for(s in seq(1, n)) {
  pick_m <- sample(52, size = m)
  # Populate the samples data frame. First, populate the cards.
  samples[s, 1:m] <- deck$cards[pick_m]
  # Then, populate the sum column.
  samples$sum[s] <-sum(deck$values[pick_m])
}
```

## Distribution of Values

```{r samples_statistics}
# Get descriptive statistics.
summary(samples)

# Create the mode function.
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Find the mode.
getmode(samples$sum)

# Find the variance.
var(samples$sum)

# Find the standard deviation of the samples.
sd(samples$sum)
```

* Measures of central tendency
    * Mean: 19.65
    * Median: 20
    * Mode: 21

* Measures of variation
    * IQR: 8
    * Variance: 28.61136
    * Standard Deviation: 5.348959

```{r samples_histogram}
ggplot(data = samples, aes(x = sum)) + 
  geom_histogram(binwidth = 1, alpha = 0.5, fill = "cyan", colour = "white") + 
  labs(x = "Sum of 3 Cards", y = "Count", title = "3 Card Sum Histogram") +
  scale_x_continuous(breaks = seq(3, 30, 3), limits=c(3, 30))
```

The distribution of sample values is normal. This is because the distribution 
of sample means (or in this case, sums) is normal for any population. This is 
given by The Central Limit Theorem.

* What range of values do you expect 90% of three-card draw sums to fall into?

90% of the draw values should fall within the 5th and 95th percentile. 
The z-scores that correspond to these percentiles are -1.645 and 1.645, 
respectively. For our estimated mean of 19.65 and our estimated standard 
deviation of 5.35, these z-scores correspond to:

z = (x - sample_mean) / SE => -1.645 = (x - 19.65) / 5.35 => x = 10.86

z = (x - sample_mean) / SE => 1.645 = (x - 19.65) / 5.35 => x = 28.45


Therefore, we expect approximately 90% of the draw values to fall in the range 
(10.86, 28.45).

* How likely are you to get a draw of at least 20 from a random draw?

Using our estimated mean and standard deviation, 20 corresponds to a z-score of:

z = (x - sample_mean) / SE => z = (20 - 19.65) / 5.35 => z = 0.065

Using this, we find that the probability that the draw value will be less than 
20 is 0.526. Therefore, the probability that the draw value will be at least 20 
is 0.474.

