---
title: "Explore Many Variables"
author: "George Traskas"
date: "2017-06-11"
output: html_document
---

## Multivariate Data

***

### Reading in Data and Loading Libraries

```{r}
pf <- read.csv("pseudo_facebook.tsv", sep = "\t")
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
```

***

### Third Qualitative Variable

```{r Third Qualitative Variable}
p1 <- ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_boxplot() +
  stat_summary(fun.y = mean, geom = "point", shape = 4)

p2 <- ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) +
  geom_line(aes(color = gender), stat = "summary", fun.y = median)

grid.arrange(p1, p2, ncol = 1)
```

### Create a new data frame, that contains information on each age and gender group.

```{r}
pf.fc_by_age_gender <- pf %>% 
  filter(!is.na(gender)) %>% 
  group_by(age, gender) %>% 
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n= n()) %>%
  arrange(age)

head(pf.fc_by_age_gender)
```

***

### Plotting Conditional Summaries

```{r Plotting Conditional Summaries}
ggplot(aes(x = age, y = median_friend_count),
       data = pf.fc_by_age_gender) +
  geom_line(aes(color = gender))
```

***

### Thinking in Ratios

***

### Wide and Long Format
Notes: Use of reshape2 or tidyr (easier)

***

### Reshaping Data

```{r}
pf.fc_by_age_gender.wide = pf.fc_by_age_gender %>%
  select(age, gender, median_friend_count) %>%
  filter(!is.na(gender)) %>% 
  spread(gender, median_friend_count) %>% 
  mutate(ratio = female / male)

head(pf.fc_by_age_gender.wide)
```

***

### Ratio Plot

```{r Ratio Plot}
ggplot(aes(x = age, y = ratio), data = pf.fc_by_age_gender.wide) +
  geom_line() +
  geom_hline(yintercept = 1, alpha = 0.3, linetype = 2)
```

***

### Third Quantitative Variable

```{r Third Quantitative Variable}
pf$year_joined = floor(2014 - pf$tenure / 365)
```

***

### Cut a Variable

```{r Cut a Variable}
summary(pf$year_joined)
table(pf$year_joined)

pf$year_joined.bucket <- cut(pf$year_joined,
                             c(2004, 2009, 2011, 2012, 2014))
table(pf$year_joined.bucket, useNA = "ifany")
```

***

### Plotting it All Together
Notes: Create a line graph of friend_count vs. age so that each year_joined.bucket is a line tracking the median user friend_count across age.

```{r Plotting it All Together}
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary", fun.y = median)
```

***

### Plot the Grand Mean
Notes: Plotting the grand mean is a good reminder that much of the data in the sample is about members of recent cohorts.

```{r Plot the Grand Mean}
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary", fun.y = mean) +
  geom_line(stat = "summary", fun.y = mean, linetype = 2)
```

***

### Friending Rate
Notes: Max 417 is definetely an outlier since 3rd quantile is only about 0.57.

```{r Friending Rate}
with(subset(pf, tenure >= 1), summary(friend_count / tenure))
```

***

### Friendships Initiated
Notes: Create a line graph of mean of friendships_initiated per day (of tenure) vs. tenure colored by year_joined.bucket.

What is the median friend rate? 0.2205

What is the maximum friend rate? 417

```{r Friendships Initiated}
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary", fun.y = mean)
```

Notes: Users with more tenure typically initiate less friendships.

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

p3 = ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

p4 = ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

p5 = ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

p6 = ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

p7 = ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))

grid.arrange(p3, p4, p5, p6, p7, ncol = 1)
```

***

### Sean's NFL Fan Sentiment Study
Notes: Bias Variance Trade Off

***

### Introducing the Yogurt Data Set

***

### Histograms Revisited

```{r Histograms Revisited}
yo = read.csv("yogurt.csv")

# Change the id from an int to a factor
yo$id = factor(yo$id)
str(yo)

p8 = ggplot(aes(price), data = yo) +
  geom_histogram(fill = "orange")

p9 = ggplot(aes(price), data = yo) +
  geom_histogram(fill = "orange", binwidth = 10)

grid.arrange(p8, p9, ncol = 1)
```

We notice
* Important discreetness
* For this very discreet data, this histogram is a very biased model.

***

### Number of Purchases
Notes: See discreetness with summary or by length and unique. The 3rd quantile is about the same with the maximum.

```{r Number of Purchases}
summary(yo)
unique(yo$price)
length(unique(yo$price))
table(yo$price)

yo = transform(yo, all.purchases = strawberry + blueberry +
                 pina.colada + plain + mixed.berry)
summary(yo$all.purchases)
```

***

### Prices over Time
Notes: The modal or most common prices seem to be increasing over time. We also see some lower prices scattered about the graph. These may be due to sales or buyers using coupons that bring the prices down.

```{r Prices over Time}
ggplot(aes(all.purchases), data = yo) +
  geom_histogram(fill = "cyan", binwidth = 1)

ggplot(aes(time, price), data = yo) +
  geom_jitter(alpha = 1/4, shape = 21, fill = "brown")
```

***

### Sampling Observations
Notes: Take a closer look applying sub-sampling.

***

### Looking at Samples of Households
Notes: x %in% y returns a logical (boolean) vector the same length as x that says whether each entry in x appears in y. That is, for each entry in x, it checks to see whether it is in y. Use the pch or shape parameter to specify the symbol when plotting points. Using the size parameter, we add more detail to the plot.

```{r Looking at Sample of Households}
# Set the seed for reproducible results
set.seed(4230)
sample.ids = sample(levels(yo$id), 16)

ggplot(aes(time, price), data = subset(yo, id %in% sample.ids)) +
  facet_wrap(~ id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)
```

***

### The Limits of Cross Sectional Data
Notes: The general idea is that we have observations over time, we can face by the primary unit, case or individual in the data set. For yogurt data, it was the households we were faceting over. The same plot cannot be generated with the pseudo Facebook data set, since we don't have data on our sample of users over time.

***

### Many Variables

***

### Scatterplot Matrix
Notes: If the plot takes a long time to render or if you want to see some of the scatterplot matrix, then only examine a smaller number of variables. You can use the following code or select fewer variables. We recommend including gender (the 6th variable)! 

pf_subset <- pf[ , c(2:7)] 

You may also find that variable labels are on the outer edges of the scatterplot matrix, rather than on the diagonal. If you want labels in the diagonal, you can set the axisLabels = 'internal' argument in your ggpairs command.

```{r}
# install.packages('GGally')
library(GGally)
theme_set(theme_minimal(10))
# Set the seed for reproducible results
set.seed(1836)

pf_subset = pf[, c(2:15)]
names(pf_subset)
ggpairs(pf_subset)
```

* Scatterplots are below the diagonal, and categorical variables, like gender, create faceted histograms.

***

### Even More Variables
Notes: Micro-array/Gene Expression Data

```{r}
nci = read.table("nci.tsv")

# Change the colnames to produce a nicer plot
colnames(nci) = c(1:64)
```

***

### Heat Maps
Notes:

```{r}
library(reshape2)
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```

***

### Analyzing Three or More Variables
Reflection
* Simple extensions to scatter plots and plots of conditional summaries
* Scatter plot matrices and heat maps
* Reshape data (long <-----> wide)

***

### Price Histograms with Facet and Color

```{r}
data("diamonds")
str(diamonds)
ggplot(aes(x = price), data = diamonds) +
  geom_histogram(aes(fill = cut)) +
  facet_wrap(~ color) +
  scale_fill_brewer(palette = "Spectral") +
  scale_x_log10()
```

### Price vs. Table Colored by Cut

```{r}
ggplot(aes(x = table, y = price, color = cut), data = diamonds) +
  geom_point() +
  scale_color_brewer(type = 'qual') +
  scale_x_continuous(limits = c(50, 80), breaks = seq(50, 80, 2))
```

### Price vs. Volume and Diamond Clarity

```{r}
diamonds$volume = diamonds$x * diamonds$y * diamonds$z
ggplot(data = subset(diamonds, volume != 0 & diamonds$volume <= quantile(diamonds$volume, 0.99)),
       aes(x = volume, y = price)) +
  geom_point(aes(color = clarity)) +
  scale_color_brewer(type = 'div') +
  scale_y_log10()
```

### Proportion Of Friendships Initiated

```{r}
pf <- transform(pf,prop_initiated = ifelse(friend_count == 0,
                                           0, friendships_initiated/friend_count))
```

### Prop_initiated Vs. Tenure

```{r}
ggplot(aes(x = tenure, y = prop_initiated),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary", fun.y = median)
```

### Smoothing prop_initiated vs. tenure

```{r}
ggplot(aes(x = tenure, y = prop_initiated),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))
```

### Largest Group Mean Prop_initiated

```{r}
lg_prop_init <- subset(pf, year_joined.bucket == '(2012,2014]')
summary(lg_prop_init$prop_initiated)
```

### Price/Carat Binned, Faceted, & Colored

```{r}
ggplot(data = diamonds, aes(x = cut, y = price/carat, color = color)) +
  geom_jitter(alpha = 0.3) +
  facet_wrap(~ clarity, ncol = 2) +
  scale_color_brewer(type = 'div')
```
