---
title: "Diamonds & Price Predictions"
author: "George Traskas"
date: "2017-06-13"
output: html_document
---

### Load Data

```{r}
library(ggplot2)
data("diamonds")
```

***

### Scatterplot Review

```{r Scatterplot Review}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(fill = "orange", color = "black", shape = 21) +
  coord_cartesian(xlim = c(0, quantile(diamonds$carat, 0.99)),
  ylim = c(0, quantile(diamonds$price, 0.99)))
```

***

### Price and Carat Relationship
Seems like an exponential relationship. The variance also increases as carat size increases.

```{r}
# A linear fit model doesn't predict the relationship.
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(fill = "orange", color = "black", shape = 21) +
  stat_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, quantile(diamonds$carat, 0.99)),
  ylim = c(0, quantile(diamonds$price, 0.99)))
```

***

### ggpairs Function

```{r ggpairs Function}
# install these if necessary
# install.packages('GGally')
# install.packages('scales')
# install.packages('memisc')
# install.packages('lattice')
# install.packages('MASS')
# install.packages('car')
# install.packages('reshape')
# install.packages('plyr')

# load the ggplot graphics package and the others
library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(gridExtra)

# sample 10,000 diamonds from the data set
set.seed(20022012)
diamond_samp <- diamonds[sample(1:length(diamonds$price), 10000), ]
ggpairs(diamond_samp, 
        lower = list(continuous = wrap("points", shape = I('.'))), 
        upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

What are some things you notice in the ggpairs output?
* Might be a relationship between price-color and price-clarity
* Strong relationship between price-carat
* Weight = f(volume) = f(x*y*z) ==> Interested in cubic root of carat

***

### The Demand of Diamonds
Notes: https://www.r-statistics.com/2013/05/log-transformations-for-skewed-and-wide-distributions-from-practical-data-science-with-r/

```{r The Demand of Diamonds}
p1 = ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 100, fill = "cyan", color = "white")
p2 = ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 0.01, fill = "orange", color = "white") +
  scale_x_log10()
grid.arrange(p1, p2)
```

***

### Connecting Demand and Price Distributions
* First histogram is heavily skewed.
* Second histogram of log10 price is kind of a normal bimodal distribution, showing two classes of poor and rich buyers.

***

### Scatterplot Transformation
Notes: The prices now look less dispersed with log10 price but we can do better transforming the carat value too.

```{r Scatterplot Transformation}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(fill = "orange", color = "black", shape = 21) +
  scale_y_log10()
```


### Create a new function to transform the carat variable

```{r cuberoot transformation}
cuberoot_trans = function() trans_new('cuberoot', transform = function(x) x^(1/3),
                                      inverse = function(x) x^3)
```

#### Use the cuberoot_trans function
Notes: With the transformation of carat and price, the relationship looks linear.

```{r Use cuberoot_trans}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point() + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Overplotting Revisited

```{r Sort and Head Tables}
head(sort(table(diamonds$carat), decreasing = TRUE))
head(sort(table(diamonds$price), decreasing = TRUE))
```


```{r Overplotting Revisited}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point(alpha = .5, size = 3/4, position = "jitter") + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Other Qualitative Factors
Notes: Clarity, Cut and Color

***

### Price vs. Carat and Clarity

Alter the code below.
```{r Price vs. Carat and Clarity}
# install and load the RColorBrewer package
# install.packages('RColorBrewer')
library(RColorBrewer)

ggplot(aes(x = carat, y = price, color = clarity), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Clarity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
    breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
    breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Clarity')
```

***

### Clarity and Price
* Diamonds with lower clarity are almost always cheaper than diamonds with better clarity.

***

### Price vs. Carat and Cut

Alter the code below.
```{r Price vs. Carat and Cut}
ggplot(aes(x = carat, y = price, color = cut), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Cut', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Cut')
```

***

### Cut and Price
* We don't see much variation on cut. Most of the diamonds in the data are ideal cut anyway.

***

### Price vs. Carat and Color

Alter the code below.
```{r Price vs. Carat and Color}
ggplot(aes(x = carat, y = price, color = color), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = F,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Color')
```

***

### Color and Price
* Color does seem to explain some of the variance in price just like we saw with clarity.

***

### Linear Models in R
Notes: Price is the outcome and carat is the predictor variable. We used our domain knowledge of diamonds and carat weight to take the cube root of carat weight (volume).

* Linear model is described by log(price)~carat^(1/3)

***

### Building the Linear Model
Notes: http://data.princeton.edu/R/linearModels.html

```{r Building the Linear Model}
m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamonds)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5, sdigits = 3)
```

Notice how adding cut to our model does not help explain much of the variance
in the price of diamonds. This fits with out exploration earlier.

***

### Model Problems
Data is from 2008, so we need to account for:
* Inflation
* 2008 global financial crisis
* Diamond Market in China is growing rapidly
* Uneven price increase across diferrent carat weight

***

### A Bigger, Better Data Set

```{r A Bigger, Better Data Set}
# install.package('bitops')
# install.packages('RCurl')
# library('bitops')
# library('RCurl')
# 
# diamondsurl = getBinaryURL("https://raw.github.com/solomonm/diamonds-data/master/BigDiamonds.Rda")
# load(rawConnection(diamondsurl))
load("BigDiamonds.Rda")
```

The code used to obtain the data is available here:
https://github.com/solomonm/diamonds-data

## Building a Model Using the Big Diamonds Data Set

```{r Building a Model Using the Big Diamonds Data Set}
diamondsbig$logprice = log(diamondsbig$price)
m1 <- lm(logprice ~ I(carat^(1/3)),
         data = diamondsbig[diamondsbig$price < 10000 &
                              diamondsbig$cert == "GIA",])
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5, sdigits = 3)
```

***

## Predictions

Example Diamond from BlueNile:
Round 1.00 Very Good I VS1 $5,601

```{r}
#Be sure you???ve loaded the library memisc and have m5 saved as an object in your workspace.
thisDiamond = data.frame(carat = 1.00, cut = "V.Good",
                         color = "I", clarity="VS1")
modelEstimate = predict(m5, newdata = thisDiamond,
                        interval="prediction", level = .95)
# We need to exponentiate the result of our model since we took the log of price.
exp(modelEstimate)
```

Evaluate how well the model predicts the BlueNile diamond's price. Think about the fitted point estimate as well as the 95% CI.

* The prediction interval here may be slightly conservative, as the model errors are heteroskedastic over carat (and hence price) even after our log and cube-root transformations.

* The results yield an expected value for price given the characteristics of our diamond and the upper and lower bounds of a 95% confidence level. The predicted value is a touch lower than the real one, though it is by no means outside of 95% CI. BlueNile has by most accounts a better reputation than diamondSE.info however. And reputation is worth a lot in a business that relies on easy to forge certificates in which the non-expert can be easily fooled. So, while this model might give you a sense of whether your diamond is a ripoff against diamondSE.info diamonds, it's not clear that diamondSE.info should be regarded as the universal source of truth over whether the price of a diamond is reasonable. Nonetheless, to have the expected price and diamondSE.info with a 95% interval, is a lot more information than we had about the price we should be willing to pay for a diamond before we started this study.

***
